\chapter{Benchmarking}

\section{Main question}

How good is your BLAST-hit? (is it a real homology o no?)

It depends on similarity.

\subsection{The confusion table}

Match hit/no hit of your data with the ones in the database in order to evaluate
the goodness of the predictor (sensitivity, specificity, accuracy, coverage,
recall).

If the hit is not in the database, we don't know how to consider that hit/no hit
since we can't match it with anything.

All positives: TP + FN
All negatives: TN + FP
All positive predictions: TP + FP

\subsection{Tradeoff}

It is easy to predict all know cases (low threshold) and also not to make any
errors (high threshold). We need to optimize false positive (fist case) and
false negative (second case).

\subsection{Measures}

\subsubsection{Sensitivity (Recall)}

$Sensitivity/Coverage/Recall = True positive rate = Sn = \frac{TP}{TP + FN}$

\subsubsection{Specificity}

$Specificity = 1 - Error = Sp = \frac{TN}{TN + FP}$

$Error = 1 - Sp = Err = \frac{FP}{TN + FP}$

\subsubsection{Precision}

$Precision = Positive predicted value = PPV = \frac{TP}{TP + FP}$

$Error per Query = (1-PPV) = EPQ = \frac{FP}{TP + FP}$

\subsubsection{Accuracy}

$Accuracy = \frac{TP + TN}{TP + FN + TN + FP}$

$Balanced Accuracy = \frac{Sn + Sp}{2} = \frac{TP}{2 \cdot (TP + FN)} +
\frac{TN}{2 \cdot (TN + FP)}$

\subsubsection{Matthews correlation coefficient}

$Matthews correlation coefficient$

\subsection{Different plots}

Receiver-Operator Characteristics (ROC)

Coverage vs Specificity: coverage vs specificity

Precision/Recall: precision vs recall

Coverage/EPQ: coverage vs error per query.

\section{Paper questions}

\begin{enumerate}
\item The advantages of using EPQ instead of FPR is that we have TP instead of
TN, because we don't know which are the TN;
\item Three classes of methods: sequence-sequence, sequence-profile,
profile-profile. They perform differently: better coverage for profile-profile
methods. Also, p-p performs better because it does (not?) uses structures;
\item The reference database was CATH. It was a fine choice since the golden
standard uses structure-structure, so we have a good way to evaluate how our
method performs;
\item In trans-membrane proteins could be better to look for sequence-sequence
instead of profile-profile.
\end{enumerate}

Why is it important to detect homologous? Because if we detect homologous, we
know that the two subjects derive from a common anchestor.
