\chapter{From DNA to Protein: How Cells Read the Genome}

\section{From DNA to RNA}

Transcription and translation are the means by which cells read out, or
express, their genetic instructions - their genes. Many identical RNA copies
can be made from the same gene, and each RNA molecule can direct
the synthesis of many identical protein molecules. Because each cell
contains only one or two copies of any particular gene, this successive
amplification enables cells to rapidly synthesize large amounts of protein
whenever necessary. At the same time, each gene can be transcribed
and translated with a different efficiency, providing the cell with a way
to make vast quantities of some proteins and tiny quantities of others.

\subsection{Portions of DNA sequence Are Transcribed into RNA}

The first step a cell takes in reading out one of its many thousands of
genes is to copy the nucleotide sequence of that gene into RNA. The process
is called transcription because the information, though copied into
another chemical form, is still written in essentially the same language -
the language of nucleotides. Like DNA, RNA is a linear polymer made of
four different types of nucleotide subunits linked together by phosphodiester
bonds. It differs from DNA chemically in two respects:
(1) the nucleotides in RNA are ribonucleotides—that is, they contain the
sugar ribose (hence the name ribonucleic acid) rather than deoxyribose;
(2) although, like DNA, RNA contains the bases adenine (A), guanine (G),
and cytosine (C), it contains uracil (U) instead of the thymine (T) found
in DNA.

Whereas DNA always occurs in cells as a
double-stranded helix, RNA is single-stranded. This difference has important
functional consequences. Because an RNA chain is single-stranded,
it can fold up into a variety of shapes, just as a polypeptide chain folds
up to form the final shape of a protein; double-stranded DNA cannot fold in this fashion.

\subsection{Transcription produces RNA Complementary to One strand of DNA}

All of the RNA in a cell is made by transcription, a process that has certain
similarities to DNA replication.
The RNA chain produced by transcription - the transcript - is therefore
elongated one nucleotide at a time and has a nucleotide sequence exactly
complementary to the strand of DNA used as the template.

Transcription, however, differs from DNA replication in several crucial
features. Unlike a newly formed DNA strand, the RNA strand does not
remain hydrogen-bonded to the DNA template strand.

The enzymes that carry out transcription are called RNA polymerases.
RNA polymerases catalyze the formation of the phosphodiester bonds that
link the nucleotides together and form the sugar-phosphate
backbone of the RNA chain.

The almost immediate release of the RNA strand from the DNA as it is
synthesized means that many RNA copies can be made from the same
gene in a relatively short time; the synthesis of the next RNA is usually
started before the first RNA has been completed.

Although RNA polymerase catalyzes essentially the same chemical reaction
as DNA polymerase, there are some important differences between
the two enzymes. First, and most obviously, RNA polymerase catalyzes
the linkage of ribonucleotides, not deoxyribonucleotides. Second, unlike
the DNA polymerase involved in DNA replication, RNA polymerases can
start an RNA chain without a primer. This difference may exist because
transcription need not be as accurate as DNA replication; unlike DNA,
RNA is not used as the permanent storage form of genetic information in
cells, so mistakes in RNA transcripts have relatively minor consequences.

\subsection{Several Types of RNA Are produced in Cells}

The vast majority of genes carried in a cell’s DNA specify the amino
acid sequence of proteins, and the RNA molecules that are copied from
these genes (and that ultimately direct the synthesis of proteins) are collectively
called messenger RNA (mRNA). In eucaryotes, each mRNA
typically carries information transcribed from just one gene, coding for a
single protein; in bacteria, a set of adjacent genes is often transcribed as
a single mRNA that therefore carries the information for several different
proteins.

As we shall see in later sections of this chapter, these nonmessenger RNAs,
like proteins, serve as regulatory, structural, and enzymatic components
of cells, and they play key parts in translating the genetic message into
protein. Ribosomal RNA (rRNA) forms the core of the ribosomes, on which
mRNA is translated into protein, and transfer RNA (tRNA) forms the adaptors
that select amino acids and hold them in place on a ribosome for
their incorporation into protein. Other small RNAs, called microRNAs
(miRNAs), serve as key regulators of eucaryotic gene expression.

In the broadest sense, the term gene expression refers to the process
by which the information encoded in a DNA sequence is translated into
a product that has some effect on a cell or organism. In cases where
the final product of the gene is a protein, gene expression includes both
transcription and translation. When an RNA molecule is the gene’s final
product, however, gene expression does not require translation.

\subsection{Signals in DNA Tell RNA polymerase Where to start and Finish}

To begin transcription, RNA polymerase must be able to recognize the start of
a gene and bind firmly to the DNA at this site.

When an RNA polymerase collides randomly with a piece of DNA,
it sticks weakly to the double helix and then slides rapidly along. The
enzyme latches on tightly only after it has encountered a region called a
promoter, which contains a specific sequence of nucleotides indicating
the starting point for RNA synthesis. Once the RNA polymerase reaches
the promoter and binds tightly to the DNA, it opens up the double helix
immediately in front of it to expose the nucleotides on each strand of a
short stretch of DNA. One of the two exposed DNA strands
then acts as a template for complementary base-pairing with incoming
ribonucleotides, two of which are joined together by the polymerase to
begin the RNA chain. Chain elongation then continues until the enzyme
encounters a second signal in the DNA, the terminator (or stop site),
where the polymerase halts and releases both the DNA template and the
newly made RNA chain.

A subunit of the bacterial polymerase, called sigma ($\sigma$) factor, is primarily
responsible for recognizing the promoter sequence on DNA. Once the
polymerase has latched onto the DNA at this site and has synthesized
approximately 10 nucleotides of RNA, the sigma factor disengages, enabling
the polymerase to move forward and continue transcribing without it.

The promoter is asymmetrical and binds the polymerase in only one orientation; thus, once properly
positioned on a promoter, the RNA polymerase has no option but to tran-
scribe the appropriate DNA strand, since transcription can proceed only
in the 5`-to-3` direction.

\subsection{Initiation of eucaryotic gene Transcription is a Complex process}

Many of the principles outlined above for bacterial transcription also
apply to eucaryotes. However, transcription initiation in eucaryotes differs
in several important ways from that in bacteria:

\begin{itemize}
\item The first difference lies in the RNA polymerases themselves. While
bacteria contain a single type of RNA polymerase, eucaryotic
cells have three - RNA polymerase I, RNA polymerase II, and RNA
polymerase III. These polymerases are responsible for transcrib-
ing different types of genes. RNA polymerases I and III transcribe
the genes encoding transfer RNA, ribosomal RNA, and small RNAs
that play structural and catalytic roles in the cell.
RNA polymerase II transcribes the vast majority of eucaryotic genes,
including all those that encode proteins.
\item A second difference is that whereas the bacterial RNA polymerase
(along with its sigma subunit) is able to initiate transcription
on its own, eucaryotic RNA polymerases require the assistance of
a large set of accessory proteins. Principal among these are the
general transcription factors, which must assemble at each pro-
moter along with the polymerase before the polymerase can begin
transcription.
\item A third distinctive feature of transcription in eucaryotes is that
the mechanisms that control its initiation are much more elaborate
than those in prokaryotes. In bacteria, genes tend to lie very close to one another
in the DNA, with only very short lengths of nontranscribed DNA
between them. But in plants and animals, including humans, individual
genes are spread out along the DNA, with stretches of up
to 100,000 nucleotide pairs between one gene and the next. This
architecture allows a single gene to be controlled by an almost
unlimited number of regulatory sequences scattered along the
DNA, and enables eucaryotes to engage in more complex forms of
transcriptional regulation than bacteria do.
\item Last but not least, eucaryotic transcription initiation must take into
account the packing of DNA into nucleosomes and more compact
forms of chromatin structure.
\end{itemize}

\subsection{Eucaryotic RNA polymerase Requires general Transcription Factors}

The initial finding that, unlike bacterial RNA polymerase, purified eucaryotic
RNA polymerase II could not on its own initiate transcription in vitro
led to the discovery and purification of the general transcription factors.
These accessory proteins assemble on the promoter, where they position
the RNA polymerase, pull apart the double helix to expose the template
strand, and launch the RNA polymerase, to begin transcribing.

The assembly process typically
begins with the binding of the general transcription factor TFIID to a short
double-helical DNA sequence primarily composed of T and A nucleotides;
because of its composition, this sequence is known as the TATA sequence,
or TATA box. Upon binding to DNA, TFIID causes a dramatic local distortion in the DNA,
which helps to serve as a landmark for the
subsequent assembly of other proteins at the promoter. The TATA box is
a key component of many promoters used by RNA polymerase II, and it
is typically located 25 nucleotides upstream from the transcription start
site. Once the first general transcription factor has bound to this DNA site,
the other factors are assembled, along with RNA polymerase II, to form a
complete transcription initiation complex.


\subsection{Eucaryotic RNAs Are Transcribed and processed simultaneously in the Nucleus}

Bacterial DNA lies directly exposed to the cytoplasm,
which contains the ribosomes on which protein synthesis takes place.
As mRNA molecules in bacteria are transcribed, ribosomes immediately
attach to the free 5` end of the RNA transcript and protein synthesis
starts.

In eucaryotic cells, by contrast, DNA is enclosed within the nucleus.
Transcription takes place in the nucleus, but protein synthesis takes place
on ribosomes in the cytoplasm. So, before a eucaryotic mRNA can be
translated, it must be transported out of the nucleus through small pores
in the nuclear envelope. Before a eucaryotic RNA exits the
nucleus, however, it must go through several different RNA processing
steps. These reactions take place as the RNA is being transcribed. The
enzymes responsible for RNA processing ride on the ‘tail’ of the eucaryo-
tic RNA polymerase as it transcribes an RNA, processing the transcript as
it emerges from the RNA polymerase.

Depending on which type of RNA is being produced - an mRNA or some
other type - the transcripts are processed in various ways before leaving
the nucleus. Two processing steps that occur only on transcripts destined
to become mRNA molecules are capping and polyadenylation:

\begin{enumerate}
\item RNA capping involves a modification of the 5` end of the mRNA
transcript, the end that is synthesized first during transcription. The
RNA is capped by the addition of an atypical nucleotide - a guanine
(G) nucleotide with a methyl group attached. This capping occurs
after the RNA polymerase has produced about 25 nucleotides of
RNA, long before it has completed transcribing the whole gene.
\item Polyadenylation provides newly transcribed mRNAs with a special
structure at their 3` ends. In contrast with bacteria, where the 3`
end of an mRNA is simply the end of the chain synthesized by the
RNA polymerase, the 3` ends of eucaryotic RNAs are first trimmed
by an enzyme that cuts the RNA chain at a particular sequence of
nucleotides and are then finished off by a second enzyme that adds
a series of repeated adenine (A) nucleotides to the cut end. This
poly-A tail is generally a few hundred nucleotides long.
\end{enumerate}

\subsection{Eucaryotic genes Are interrupted by Noncoding sequences}

Most eucaryotic RNAs have to undergo an additional processing step
before they are functional.

Most eucaryotic genes, in contrast, have their coding sequences interrupted
by long, noncoding intervening sequences, called introns. The scattered pieces
of coding sequences, or expressed sequences, called exons, are usually shorter than
the introns, and the coding portion of a eucaryotic gene is often only a
small fraction of the total length of the gene.

\subsection{Introns Are Removed by RNA splicing}

To produce an mRNA in a eucaryotic cell, the entire length of the gene,
introns as well as exons, is transcribed into RNA. After capping, and as
the RNA polymerase continues to transcribe the gene, the process of RNA
splicing begins, in which the intron sequences are removed from the
newly synthesized RNA and the exons are stitched together. Each transcript
ultimately receives a poly-A tail.
Once a transcript has been spliced and its 5` and
3` ends have been modified, the RNA is a functional mRNA molecule that
can now leave the nucleus and be translated into protein.

Unlike the other steps of mRNA production we have discussed, RNA
splicing is carried out largely by RNA molecules instead of proteins. RNA
molecules recognize intron–exon boundaries (through complementary
base-pairing) and participate intimately in the chemistry of splicing.
These RNA molecules, called small nuclear RNAs (snRNAs), are packaged
with additional proteins to form small nuclear ribonucleoprotein
particles (snRNPs, pronounced “snurps”). The snRNPs form the core of
the spliceosome, the large assembly of RNA and protein molecules that
carries out RNA splicing in the cell.

The intron-exon type of gene arrangement in eucaryotes at first seems
wasteful, but it does have positive consequences. First, the transcripts of
many eucaryotic genes can be spliced in different ways, each of which
can produce a distinct protein. Such alternative splicing thereby allows
many different proteins to be produced from the same gene.
Thus RNA splicing enables eucaryotes to increase the already enormous
coding potential of their genomes.

The intron-exon structure of genes is thought to have speeded up the emergence of new and
useful proteins. The inclusion of long introns makes genetic recombination
between exons of different genes much more likely. This means that
genes for new proteins could have evolved quite rapidly by the combination
of parts of preexisting genes, a mechanism resembling the assembly
of a new type of machine from a kit of preexisting functional components.
Indeed, many proteins in present-day cells resemble patchworks
composed from a common set of protein pieces, called protein domains.

\subsection{Mature eucaryotic mRNAs Are selectively exported from the Nucleus}

We have seen how eucaryotic mRNA synthesis and processing takes
place in an orderly fashion within the cell nucleus. However, these events
create a special problem for eucaryotic cells: of the total mRNA that is
synthesized, only a small fraction -t he mature mRNA - is useful to the
cell. The remaining RNA fragments-excised introns, broken RNAs, and
aberrantly spliced transcripts - are not only useless but could be dangerous
to the cell if not destroyed. How, then, does the cell distinguish
between the relatively rare mature mRNA molecules it needs to keep and
the overwhelming amount of debris generated by RNA processing?

The answer is that the transport of mRNA from the nucleus to the
cytoplasm, where it is translated into protein, is highly selective: only
correctly processed RNAs are allowed to pass. This reliance on proper
processing for RNA transport is mediated by the nuclear pore complex,
which recognizes and exports only completed mRNAs. These aqueous
pores connect the nucleoplasm with the cytosol.
They act as gates that control which macromolecules can
enter or leave the nucleus.

The ‘waste RNAs’ that remain behind in the nucleus are degraded, and the building
blocks are reused for transcription.

\subsection{mRNA Molecules Are eventually Degraded by the Cell}

Because a single mRNA molecule can be translated many times, the length of time
that a mature mRNA molecule persists in the cell affects the amount of protein it produces.
Each mRNA molecule is eventually degraded into nucleotides by RNases, but the lifetimes
of mRNA molecules differ considerably - epending on the nucleotide
sequence of the mRNA and the type of cell in which the mRNA is produced.

These different lifetimes are in part controlled by nucleotide sequences
that are present in the mRNA itself, most often in the portion of RNA
called the 3` untranslated region, that lies between the 3` end of the cod-
ing sequence and the poly-A tail.

\subsection{The earliest Cells May have had introns in Their genes}

The process of transcription is universal: all cells use RNA polymerase,
coupled with complementary base-pairing, to synthesize RNA from DNA.

\section{From RNA to Protein}

How is the information in a linear
sequence of nucleotides in RNA translated into the linear sequence of a
chemically quite different set of subunits—the amino acids in proteins?

\subsection{An mRNA sequence is Decoded in sets of Three Nucleotides}

The conversion of the information in RNA into protein represents a translation
of the information into another language that uses
quite different symbols. Because there are only 4 different nucleotides in
mRNA but 20 different types of amino acids in a protein, this translation
cannot be accounted for by a direct one-to-one correspondence between
a nucleotide in RNA and an amino acid in protein. The rules by which the
nucleotide sequence of a gene, through the medium of mRNA, is translated
into the amino acid sequence of a protein are known as the genetic
code.

Because RNA is a linear polymer made of four different 
nucleotides, there are thus $4 \cdot 4 \cdot 4 = 64$ possible combinations of
three nucleotides. However, only 20 different amino acids are commonly
found in proteins. Either some nucleotide triplets are never used, or
the code is redundant and some amino acids are
specified by more than one triplet. The second possibility is, in fact, correct.
Each group of three consecutive nucleotides in RNA is called a codon,
and each specifies one amino acid.

In principle, an RNA sequence can be translated in any one of three different
reading frames, depending on where the decoding process begins. However,
only one of the three possible reading frames in an mRNA specifies the correct protein.

\subsection{tRNA Molecules Match Amino Acids to Codons in mRNA}

The codons in an mRNA molecule do not directly recognize the amino
acids they specify: the group of three nucleotides does not, for example,
bind directly to the amino acid. Rather, the translation of mRNA into
protein depends on adaptor molecules that can recognize and bind to a
codon at one site on their surface and to an amino acid at another site.
These adaptors consist of a set of small RNA molecules known as transfer RNAs (tRNAs).

We saw earlier that an RNA molecule will generally fold into a three-dimensional
structure by forming base pairs between different regions
of the molecule. If the base-paired regions are sufficiently extensive, they
will form a double-helical structure, like that of double-stranded DNA.
The tRNA molecule provides a striking example of this. Four short segments
of the folded tRNA are double-helical, producing a molecule that
looks like a cloverleaf when drawn schematically.
Two regions of unpaired nucleotides situated at either end of the L-shaped
molecule are crucial to the function of tRNA in protein synthesis. One
of these regions forms the anticodon, a set of three consecutive nucle-
otides that through base-pairing bind the complementary codon in an
mRNA molecule. The other is a short single-stranded region at the 3` end
of the molecule; this is the site where the amino acid that matches the
codon is attached to the tRNA.

Some amino acids have more than one tRNA and some tRNAs are constructed so that
they require accurate base-pairing only at the fi rst two positions of the
codon and can tolerate a mismatch (or wobble) at the third position. This
wobble base-pairing explains why so many of the alternative codons
for an amino acid differ only in their third nucleotide.

\subsection{Specific enzymes Couple tRNAs to the Correct Amino Acid}

We have seen that in order to read the genetic code in DNA, cells make
many different tRNAs. We now must consider how each tRNA molecule
becomes charged—linked to the one amino acid in 20 that is its right partner.
Recognition and attachment of the correct amino acid depend on
enzymes called aminoacyl-tRNA synthetases, which covalently couple
each amino acid to its appropriate set of tRNA molecules. In most organisms,
there is a different synthetase enzyme for each amino acid. The
synthetases are equal in importance to the tRNAs in the decoding process,
because it is the combined action of synthetases and tRNAs that
allows each codon in the mRNA molecule to associate with its proper
amino acid.

The synthetase-catalyzed reaction that attaches the amino acid to the 3`
end of the tRNA is one of many reactions coupled to the energy-releasing
hydrolysis of ATP, and it produces a high-energy bond
between the charged tRNA and the amino acid.

\subsection{The RNA Message is Decoded on Ribosomes}

The recognition of a codon by the anticodon on a tRNA molecule depends
on the same type of complementary base-pairing used in DNA replication
and transcription. However, accurate and rapid translation of mRNA
into protein requires a large molecular machine that moves along the
mRNA, captures complementary tRNA molecules, holds them in position,
and covalently links the amino acids that they carry so as to form a
protein chain. This protein-manufacturing machine is the ribosome - a
large complex made from more than 50 different proteins (the ribosomal
proteins) and several RNA molecules called ribosomal RNAs (rRNAs). A
typical living cell contains millions of ribosomes in its cytoplasm.

Eucaryotic and procaryotic ribosomes are very similar in structure and
in function. Both are composed of one large and one small subunit that
fit together to form a complete ribosome with a mass of several million daltons.
The small subunit matches the tRNAs to
the codons of the mRNA, while the large subunit catalyzes the forma-
tion of the peptide bonds that covalently link the amino acids together
into a polypeptide chain.

How does the ribosome choreograph all the movements required for
translation? Each ribosome contains a binding site for an mRNA mol-
ecule and three binding sites for tRNA molecules, called the A-site, the
P-site, and the E-site.

\subsection{The Ribosome is a Ribozyme}

Not only are the three tRNA-binding sites (the A-, P-, and E-sites) on the
ribosome formed primarily by the rRNAs, but the catalytic site for peptide
bond formation is formed by the 23S RNA of the large subunit.

RNA molecules that possess catalytic activity are called ribozymes.

\subsection{Codons in mRNA signal Where to start and to stop protein synthesis}

The translation of an mRNA begins with the codon AUG, and a special
tRNA is required to initiate translation. This initiator tRNA always carries
the amino acid methionine (or a modified form of methionine, formylmethionine,
in bacteria) so that newly made proteins all have methionine
as the first amino acid at their N-terminal end, the end of a protein that
is synthesized first. This methionine is usually removed later by a specific
protease. The initiator tRNA is distinct from the tRNA that normally carries methionine.
In eucaryotes, the initiator tRNA, coupled to methionine, is first loaded into
a small ribosomal subunit, along with additional proteins called translation
initiation factors. Of all the charged tRNAs in the cell,
only the charged initiator tRNA is capable of binding tightly to the P-site
of the small ribosomal subunit. Next, the loaded ribosomal subunit binds
to the 5` end of an mRNA molecule, which is signaled by the cap that is
present on eucaryotic mRNA. The small ribosomal sub-
unit then moves forward (5` to 3`) along the mRNA searching for the first
AUG. When this AUG is encountered, several initiation factors dissociate
from the small ribosomal subunit to make way for the large ribosomal
subunit to assemble and complete the ribosome. Because the initiator
tRNA is bound to the P-site, protein synthesis is ready to begin with the
addition of the next charged tRNA to the A-site.

The end of the protein-coding message in both procaryotes and eucaryotes
is signaled by the presence of one of several codons called stop codons.
These special codons - UAA, UAG, and UGA - are not
recognized by a tRNA and do not specify an amino acid, but instead signal
to the ribosome to stop translation. Proteins known as release factors bind
to any stop codon that reaches the A-site on the ribosome, and this binding
alters the activity of the peptidyl transferase in the ribosome, causing
it to catalyze the addition of a water molecule instead of an amino acid
to the peptidyl-tRNA. This reaction frees the carboxyl end
of the polypeptide chain from its attachment to a tRNA molecule, and
because this is the only attachment that holds the growing polypeptide to
the ribosome, the completed protein chain is immediately released into
the cytosol. The ribosome releases the mRNA and dissociates into its two
separate subunits, which can then assemble on another mRNA molecule
to begin a new round of protein synthesis.

Many proteins can fold into their three-dimensional shape spontaneously,
and some do so as they are spun out of the
ribosome. Most proteins, however, require molecular chaperones to help
them fold correctly in the cell.

\subsection{Proteins Are Made on polyribosomes}

If the mRNA is being translated efficiently, a new ribosome hops onto the 5`
end of the mRNA molecule almost as soon as the preceding ribosome has
translated enough of the nucleotide sequence to move out of the way.
The mRNA molecules being translated are therefore usually found in
the form of polyribosomes (also known as polysomes), large cytoplasmic
assemblies made up of several ribosomes spaced as close as 80 nucleotides
apart along a single mRNA molecule.

\subsection{Inhibitors of procaryotic protein synthesis Are used as Antibiotics}

Many of our most effective antibiotics are compounds that act by inhibiting
bacterial, but not eucaryotic, protein synthesis.

\subsection{Carefully Controlled protein breakdown helps Regulate the Amount of each protein in a Cell}

After a protein is released from the ribosome, it becomes subject to a
number of controls by the cell. The number of copies of a protein in a cell
depends, like the human population, not only on how quickly new individuals
are made but also on how long they survive. How does the cell control these lifetimes?

Cells possess specialized pathways to enzymatically break proteins
down into their constituent amino acids (a process termed proteolysis).
The enzymes that degrade proteins, first to short peptides and finally to
individual amino acids, are known collectively as proteases. Proteases
act by cutting (hydrolyzing) the peptide bonds between amino acids.

Although most damaged proteins are broken down in the cytosol,
important protein degradation pathways also operate in other compartments
in eucaryotic cells, such as lysosomes.
In eucaryotic cells, most proteins are broken down by machines called
proteasomes. A proteasome contains a central cylinder formed from
proteases whose active sites into an inner chamber.

How does the proteasome select which proteins in the cell should enter
the cylinder and be degraded? Proteasomes act primarily on proteins that
have been marked for destruction by the covalent attachment of a small
protein called ubiquitin. Specialized enzymes tag selected proteins with
one or more ubiquitin molecules; these ubiquitylated proteins are then
recognized and sucked into the proteasome by proteins in the stopper.

\subsection{There Are Many steps between DNA and protein}

The final concentration of a protein in
a cell therefore depends on the efficiency with which each of the many
steps is carried out. In addition, many proteins - once they leave the
ribosome - require further attention before they are useful to the cell.
We will see in the next chapter that cells have the ability to change the
concentrations of most of their proteins according to their needs.

\section{RNA and the origins of Life}

If nucleic acids are required
to direct the synthesis of proteins, and proteins are required to synthesize
nucleic acids, how could this system of interdependent components have
arisen? One view is that an RNA world existed on Earth before modern
cells appeared. According to this hypothesis, RNA - which
today serves as an intermediate between genes and proteins - both stored
genetic information and catalyzed chemical reactions in primitive cells.
Only later in evolutionary time did DNA take over as the genetic material
and proteins become the major catalysts and structural components of
cells. If this idea is correct, then the transition out of the RNA world was
never completed; as we have seen in this chapter, RNA still catalyzes
several fundamental reactions in modern cells. These RNA catalysts,
including the ribosome and RNA-splicing machinery, can thus be viewed
as molecular fossils of an earlier world.

\subsection{Life require autocatalysis}

Catalysts with this special self-promoting property, once they had arisen
by chance, would reproduce themselves and would therefore divert raw
materials from the production of other substances. In this way one can
envisage the gradual development of an increasingly complex chemical
system of organic monomers and polymers that function together to generate
more molecules of the same types, fueled by a supply of simple
raw materials in the environment. Such an autocatalytic system would
have many of the properties we think of as characteristic of living matter.

\subsection{RNA Can both store information and Catalyze Chemical Reactions}

We have seen that complementary base-pairing enables one nucleic acid
to act as a template for the formation of another. Such
complementary templating mechanisms lie at the heart of DNA replication
and transcription in modern-day cells.

But the efficient synthesis of polynucleotides by such complementary
templating mechanisms also requires catalysts to promote the polymerization
reaction. The unique potential of RNA molecules to
act both as information carriers and as catalysts is thought to have enabled
them to play the central role in the origin of life.

As we saw previously, protein enzymes are able to catalyze biochemical
reactions because they have surfaces with unique contours and chemical
properties on which a given substrate can react. In the same way,
RNA molecules, with their unique folded shapes, can serve as enzymes,
although the fact that they are constructed of only four different
subunits limits their catalytic efficiency and their range of chemical
reactions compared with proteins. Nonetheless, ribozymes can catalyze
many types of chemical reactions.

RNA, therefore, has all the properties required of a molecule that could
catalyze its own synthesis.

\subsection{RNA is Thought to predate DNA in evolution}

Evidence that RNA arose before DNA in evolution can be found in the
chemical differences between them. Ribose, like glucose and other
simple carbohydrates, is readily formed from formaldehyde
(HCHO), which is one of the principal products of experiments simulating
conditions on the primitive Earth. The sugar deoxyribose is harder to
make, and in present-day cells it is produced from ribose in a reaction
catalyzed by a protein enzyme, suggesting that ribose predates deoxyribose
in cells. Presumably, DNA appeared on the scene later, and then
proved more suited than RNA as a permanent repository of genetic information.
In particular, the deoxyribose in its sugar-phosphate backbone
makes chains of DNA chemically much more stable than chains of RNA,
so that greater lengths of DNA can be maintained without breakage.

The other differences between RNA and DNA - the double-helical structure
of DNA and the use of thymine rather than uracil - further enhance
DNA stability by making the molecule easier to repair.
Furthermore, deamination, one of the most common unwanted chemical changes occurring in
polynucleotides, is easier to detect and repair in DNA than in RNA.

As cells more closely resembling present-day cells appeared, it is
believed that many of the functions originally performed by RNA were
taken over by molecules more specifically fitted to the tasks required.
Eventually DNA took over the primary genetic function, and proteins
became the major catalysts, while RNA remained primarily as the intermediary
connecting the two.

\section{Essential concepts}

\begin{itemize}
\item The flow of genetic information in all living cells is DNA $\rightarrow$ RNA $\rightarrow$
protein. The conversion of the genetic instructions in DNA into RNAs
and proteins is termed gene expression.
\item To express the genetic information carried in DNA, the nucleotide
sequence of a gene is first transcribed into RNA. Transcription is catalyzed by the enzyme RNA polymerase. Nucleotide sequences in the
DNA molecule indicate to the RNA polymerase where to start and
stop transcribing.
\item RNA differs in several respects from DNA. It contains the sugar ribose
instead of deoxyribose and the base uracil (U) instead of thymine (T).
RNAs in cells are synthesized as single-stranded molecules, which
often fold up into precise three-dimensional shapes.
\item Cells make several different functional types of RNAs, including
messenger RNA (mRNA), which carries the instructions for making
proteins; ribosomal RNA (rRNA), which is a component of ribosomes;
and transfer RNA (tRNA), which acts as an adaptor molecule in protein synthesis.
\item Transcription begins at DNA sites called promoters. To initiate tran-
scription, eucaryotic RNA polymerases require the assembly of a
complex of general transcription factors at the promoter, whereas
bacterial RNA polymerase requires only an additional subunit, called
sigma factor.
\item In eucaryotic DNA, most genes are composed of a number of smaller
coding regions (exons) interspersed with noncoding regions (introns).
When a eucaryotic gene is transcribed from DNA into RNA, both the
exons and introns are copied.
\item Introns are removed from the RNA transcripts in the nucleus by the
process of RNA splicing. In a reaction catalyzed by small ribonucleoprotein
complexes known as snRNPs, the introns are excised from
the RNA and the exons are joined together.
\item Eucaryotic mRNAs go through several additional RNA processing
steps before they leave the nucleus, including RNA capping and polyadenylation.
These reactions, along with splicing, take place as the
RNA is being transcribed. The mature mRNA is then transported to
the cytoplasm.
\item Translation of the nucleotide sequence of mRNA into a protein takes
place in the cytoplasm on large ribonucleoprotein assemblies called
ribosomes. As the mRNA is threaded through a ribosome, its message
is translated into protein.
\item The nucleotide sequence in mRNA is read in sets of three nucleotides
(codons), each codon corresponding to one amino acid.
\item The correspondence between amino acids and codons is specified
by the genetic code. The possible combinations of the 4 different
nucleotides in RNA give 64 different codons in the genetic code. Most
amino acids are specified by more than one codon.
\item tRNA acts as an adaptor molecule in protein synthesis. Enzymes
called aminoacyl-tRNA synthetases link amino acids to their appropriate
tRNAs. Each tRNA contains a sequence of three nucleotides,
the anticodon, which matches a codon in mRNA by complementary
base-pairing between codon and anticodon.
\item Protein synthesis begins when a ribosome assembles at an initiation
codon (AUG) in mRNA, a process that is regulated by proteins
called translation initiation factors. The completed protein chain is
released from the ribosome when a stop codon (UAA, UAG, or UGA)
is reached.
\item The stepwise linking of amino acids into a polypeptide chain is catalyzed
by an rRNA molecule in the large ribosomal subunit. Thus, the
ribosome is an example of a ribozyme, an RNA molecule that can
catalyze a chemical reaction.
\item The degradation of proteins in the cell is carefully controlled. Some
proteins are degraded in the cytosol by large protein complexes
called proteasomes.
\item From our knowledge of present-day organisms and the molecules
they contain, it seems likely that living systems began with the evolution
of RNA molecules that could catalyze their own replication.
\item It has been proposed that, as cells evolved, the DNA double helix
replaced RNA as a more stable molecule for storing genetic information,
and proteins replaced RNAs as major catalytic and structural
components. However, important reactions such as peptide bond
formation are still catalyzed by RNA; these are thought to provide a
glimpse into an ancient, RNA-based world.
\end{itemize}

\chapter{Control of Gene Expression}

We need to know how the elements in the DNA sequence or the words on a list work
together to make the masterpiece.
In cell biology, the question comes down to gene expression.

\section{An overview of gene expression}

How does an individual cell specify which of its many thousands of genes
to express? This decision is an especially important problem for multicellular 
organisms because, as the animal develops, cell types such as
muscle, nerve, and blood cells become different from one another, eventually 
leading to the wide variety of cell types seen in the adult. Such
differentiation arises because cells make and accumulate different sets
of RNA and protein molecules: that is, they express different genes.

\subsection{The different Cell Types of a multicellular Organism Contain the Same DNA}

Cells have the ability to change which genes they
express without altering the nucleotide sequence of their DNA. 
If DNA were altered irreversibly during development,
the chromosomes of a differentiated cell would be incapable of guiding 
the development of the whole organism. DNA in specialized cell types
still contains the entire set of instructions needed to form a whole organism. 
The cells of an organism therefore differ not because they contain
different genes, but because they express them differently.

\subsection{Different Cell types produce different sets of proteins}

Many proteins are common to all the cells of a multicellular organism. 
These housekeeping proteins include the structural
proteins of chromosomes, RNA polymerases, DNA repair enzymes, ribosomal 
proteins, enzymes involved in glycolysis and other basic metabolic
processes, and many of the proteins that form the cytoskeleton. Each different 
cell type also produces specialized proteins that are responsible for
the cell’s distinctive properties.

\subsection{A Cell Can Change the Expression of Its Genes in Response to External Signals}

Most of the specialized cells in a multicellular organism are capable of
altering their patterns of gene expression in response to extracellular
cues. Adjustments are features of the gene expression pattern
that do not change and give each cell type its permanently distinctive
character.

\subsection{Gene Expression Can Be Regulated at many of the Step in the Pathway from DNA to RNA to Protein}

A cell can control the proteins it makes
by (1) controlling when and how often a given gene is transcribed, (2)
controlling how an RNA transcript is spliced or otherwise processed, (3)
selecting which mRNAs are exported from the nucleus to the cytosol,
(4) selectively degrading certain mRNA molecules, (5) selecting which
mRNAs are translated by ribosomes, or (6) selectively activating or inactivating 
proteins after they have been made.

For most genes, however, the control of transcription is paramount.

\section{How transcriptional switches work}

We begin with a discussion of the transcription regulators,
proteins that control gene expression at the level of transcription.

\subsection{Transcription Is Controlled by Proteins Binding to Regulatory DNA Sequences}

We saw that the promoter region of a gene
attracts the enzyme RNA polymerase and correctly orients the enzyme to
begin its task of making an RNA copy of the gene. The promoters of both
bacterial and eucaryotic genes include an initiation site, where transcription 
actually begins, and a sequence of approximately 50 nucleotides that
extends upstream from the initiation site (if one likens the direction of
transcription to the flow of a river). This region contains sites that are
required for the RNA polymerase to bind to the promoter. In addition
to the promoter, nearly all genes, whether bacterial or eucaryotic, have
regulatory DNA sequences that are used to switch the gene on or off.

Regulatory DNA sequences do not work by themselves. To have any
effect, these sequences must be recognized by proteins called transcription 
regulators, which bind to the DNA. It is the combination of a DNA
sequence and its associated protein molecules that acts as the switch
to control transcription.

Proteins that recognize a specific DNA sequence do so because the surface 
of the protein fits tightly against the special surface features of the
double helix in that region. These features will vary depending on the
nucleotide sequence, and thus different proteins will recognize different
nucleotide sequences.

The DNA-binding motifs - the homeodomain, the zinc finger, and the leucine zipper - 
are found in transcription regulators that control the expression of thousands of different
genes in virtually all eucaryotic organisms. 

\subsection{Transcription Switches Allow Cells to Respond to Changes in the Environment}

Bacteria regulate the expression of many of their genes according to the food sources
that are available in the environment. These
genes are arranged in a cluster on the chromosome and are transcribed
from a single promoter as one long mRNA molecule from which the five
proteins are translated. These five coordinately expressed genes are part of an
operon — a set of genes that are transcribed into a single mRNA.

We now understand in considerable detail how the tryptophan operon
functions. Within the promoter is a short DNA sequence (15 nucleotides
in length) that is recognized by a transcription regulator. When this protein 
binds to this nucleotide sequence, termed the operator, it blocks
access of RNA polymerase to the promoter; this prevents transcription
of the operon and production of the tryptophan-producing enzymes. The
transcription regulator is known as the tryptophan repressor, and it is controlled 
in an ingenious way: the repressor can bind to DNA only if it has
also bound several molecules of the amino acid tryptophan.

The tryptophan repressor is an allosteric protein: the
binding of tryptophan causes a subtle change in its three-dimensional
structure so that the protein can bind to the operator sequence. When
the concentration of free tryptophan in the cell drops, the repressor no
longer binds tryptophan and thus no longer binds to DNA, and the tryptophan 
operon is transcribed.

\subsection{Repressors Turn Genes Off, Activators Turn Them On}

The tryptophan repressor, as its name suggests, is a repressor protein:
in its active form, it switches genes off, or represses them. Some bacterial
transcription regulators do the opposite: they switch genes on, or activate
them. These activator proteins work on promoters that—in contrast to
the promoter for the tryptophan operator - are, on their own, only marginally 
able to bind and position RNA polymerase; they may, for example,
be recognized only poorly by the polymerase. However, these poorly
functioning promoters can be made fully functional by activator proteins
that bind to a nearby site on the DNA and contact the RNA polymerase to
help it initiate transcription. In some cases, a bacterial transcription 
regulator can repress transcription at one promoter and activate
transcription at another; whether the regulatory protein acts as an activator 
or repressor depends, in large part, on exactly where the regulatory
sequences to which it binds are located with respect to the promoter.

Like the tryptophan repressor, activator proteins often have to interact
with a second molecule to be able to bind DNA.

\subsection{An Activator and a Repressor Control the Lac Operon}

In many instances, the activity of a single promoter is controlled by two
different transcription regulators. The Lac operon in E. coli, for example,
is controlled by both the Lac repressor and the activator protein CAP. The
Lac operon encodes proteins required to import and digest the disaccharide 
lactose. In the absence of glucose, CAP switches on genes that
allow the cell to utilize alternative sources of carbon - including lactose.

\subsection{Eucaryotic Transcription Regulators Control Gene Expression from a distance}

The DNA sites to which eucaryotic gene activators bound were originally termed enhancers,
because their presence dramatically enhanced, or increased, the rate of
transcription. It was surprising to biologists when, in 1979, it was discovered 
that these activator proteins could enhance transcription even when
they are bound thousands of nucleotide pairs away from a gene’s promoter.

Many models for this ‘action at a distance’ have been proposed, but
the simplest of these seems to apply in most cases. The DNA between
the enhancer and the promoter loops out to allow eucaryotic activator 
proteins to directly influence events that take place at the promoter. 
The DNA thus acts as a tether, causing a protein bound to
an enhancer even thousands of nucleotide pairs away to interact with
the proteins in the vicinity of the promoter - including RNA polymerase
II and the general transcription factors. Often, additional proteins serve 
to link the distantly bound transcription regulators
to these proteins at the promoter; the most important is a large complex 
of proteins known as Mediator. One of the ways in
which eucaryotic activator proteins function is by aiding in the assembly
of the general transcription factors and RNA polymerase at the promoter.
Eucaryotic repressor proteins do the opposite: they decrease transcription
by preventing or sabotaging the assembly of the same protein complex.

\subsection{Packing of Promoter DNA into Nucleosomes Affects Initiation of Transcription}

In eucaryotic cells, activator and repressor proteins exploit chromatin
structure to help turn genes on and off. As we already saw, chromatin 
structure can be altered by chromatin-remodeling complexes and by
enzymes that covalently modify the histone proteins that form the core
of the nucleosome. This modification alters chromatin structure,
probably allowing greater accessibility to the underlying DNA; moreover,
the acetyl groups themselves are recognized by proteins that promote
transcription, including some of the general transcription factors.

Likewise, gene repressor proteins can modify chromatin in ways that
reduce the efficiency of transcription initiation. For example, many repressors 
attract histone deacetylases-enzymes that remove the acetyl groups
from histone tails, thereby reversing the positive effects that acetylation
has on transcription initiation.

\section{The molecular mechanisms that create specialized cell types}

Cells of multicellular organisms have evolved
this capacity to an extreme degree and in highly specialized ways to form
an organized array of differentiated cell types. In particular, once a cell in a
multicellular organism becomes committed to differentiate into a specific
cell type, the choice of fate is generally maintained through many subsequent 
cell generations. This means that the changes in gene expression,
which are often triggered by a transient signal, must be remembered. This
phenomenon of cell memory is a prerequisite for the creation of organized 
tissues and for the maintenance of stably differentiated cell types.

\subsection{Eucaryotic Genes Are Regulated by Combinations of Proteins}

The term combinatorial control refers to the way that groups of regulatory 
proteins work together to determine the expression of a single gene.
We saw a simple example of such regulation by multiple signals when
we discussed the bacterial Lac operon. In eucaryotes, the
regulatory inputs have been amplified, and a typical gene is controlled
by dozens of transcription regulators. Often, some of these
regulatory proteins are repressors and some are activators; the molecular
mechanisms by which the effects of all of these proteins are added up to
determine the final level of expression for a gene are only now beginning
to be understood.

\subsection{The Expression of different Genes Can Be Coordinated by a Single Protein}

In addition to being able to switch individual genes on and off, all
organisms—whether procaryote or eucaryote—need to coordinate the
expression of different genes. When a eucaryotic cell receives a signal to
divide, for example, a number of hitherto unexpressed genes are turned
on together to set in motion the events that lead eventually to cell division.

So how do eucaryotes coordinate gene expression? In particular, given that a eucaryotic
cell uses a committee of transcription regulators to control each of its genes, 
how can it rapidly and decisively
switch whole groups of genes on or off? The answer is that even though
control of gene expression is combinatorial, the effect of a single transcription 
regulator can still be decisive in switching any particular gene
on or off, simply by completing the combination needed to activate or
repress that gene.

An example of this style of regulation in humans is seen with the glucocorticoid 
receptor protein. In order to bind to regulatory sites in DNA
this transcription regulator must first form a complex with a molecule
of a glucocorticoid hormone

\subsection{Combinatorial Control Can Create different Cell Types}

The ability to switch many different genes on or off using just one protein
is not only useful in the day-to-day regulation of cell function. It is also
one of the means by which eucaryotic cells differentiate into particular
types of cells during embryonic development.

The conversion of one cell type (fibroblast) to another (muscle) by a single
transcription regulator emphasizes one of the most important principles
discussed in this chapter: the dramatic differences between cell types - 
such as size, shape, and function - are produced by differences in gene
expression.

\subsection{Stable Patterns of Gene Expression Can Be Transmitted to daughter Cells}

Once a cell in a multicellular organism has become differentiated into a particular cell type, it will generally
remain differentiated, and all its progeny cells will remain that same cell
type. Some highly specialized cells never divide again once they have
differentiated; for example, skeletal muscle cells and neurons. But many
other differentiated cells, such as fibroblasts, smooth muscle cells, and
liver cells (hepatocytes), will divide many times in the life of an individual. 
All of these cell types give rise only to cells like themselves when
they divide.

This preservation of cellular identity means that the changes in gene
expression that give rise to a differentiated cell must be remembered and
passed on to its daughter cells through all subsequent cell divisions.

Cells have several ways of ensuring that their daughters “remember”
what kind of cells they are supposed to be. One of the simplest is through
a positive feedback loop, where a key transcription regulator activates
transcription of its own gene in addition to that of other cell-type - specific
genes. Another way of maintaining cell type is through
the faithful propagation of a condensed chromatin structure from parent
to daughter cell.

A third way in which cells can transmit information about gene expression 
to their progeny is through DNA methylation. In vertebrate cells,
DNA methylation occurs exclusively on cytosine bases. This
covalent modification of cytosines generally turns off genes by attracting 
proteins that block gene expression. DNA methylation patterns are
passed on to progeny cells by the action of an enzyme that copies the
methylation pattern on the parent DNA strand to the daughter DNA strand
immediately after replication.

\subsection{The Formation of an Entire Organ Can Be Triggered by a Single Transcription Regulator}

We have seen that even though combinatorial control is the norm for
eucaryotic genes, a single transcription regulator can be decisive in
switching a whole set of genes on or off, and can convert one cell type
into another. So the action of just one transcription regu-
lator can produce a cascade of regulators whose combined actions lead
to the formation of an organized group of many different types of cells.
One can begin to imagine how, by repeated applications of this principle,
a complex organism is built piece by piece.

\section{Post-transcriptional controls}

We have seen that transcription regulators control gene expression by
switching on or off transcription initiation. The vast majority of genes
in all organisms are regulated in this way. But additional points of control 
can come into play later in the pathway from DNA to protein, giving
cells a further opportunity to manage the amount of gene product that
is made. These post-transcriptional controls, which operate after RNA
polymerase has bound to a gene’s promoter and started to synthesize
RNA, are crucial for the regulation of many genes.

\subsection{Riboswitches Provide An Economical Solution to Gene Regulation}

Scientists have recently discovered a number of mRNAs that can regulate their
own transcription and translation. These self-regulating mRNAs contain
riboswitches: short sequences of RNA that change their conformation
when bound to small molecules such as metabolites.
The conformational change that is driven by the binding of that molecule
can regulate gene expression.

\subsection{The untranslated Regions of mRNAs Can Control Their Translation}

Once an mRNA has been synthesized, one of the most common ways of
regulating how much of its protein product is made is to control the initiation of translation.

In eucaryotic cells, repressors can inhibit translation initiation by binding 
to specific RNA sequences in the 5` untranslated region of the mRNA
and keeping the ribosome from finding the first AUG. When conditions
change, the cell can inactivate the repressor and thereby increase translation of the mRNA.

\subsection{Small Regulatory RNAs Control the Expression of Thousands of Animal and Plant Genes}

A recent series of striking discoveries has revealed
that noncoding RNAs—those that do not direct the production of a protein 
product - are far more prevalent than previously imagined and play
unanticipated, widespread roles in regulating gene expression.

One particularly important type of noncoding RNA, found in plants and
animals, is called microRNA (miRNA). These short, regulatory RNAs control
gene expression by base-pairing with specific mRNAs and controlling
their stability and their translation.

Like other noncoding RNAs, such as tRNA and rRNA, the precursor
miRNA transcript undergoes a special type of processing to yield the
mature miRNA. This miRNA is then assembled with specialized proteins
to form an RNA-induced silencing complex (RISC). The RISC patrols the
cytoplasm, searching for mRNAs that are complementary to the miRNA
it carries. A single miRNA - as part of a RISC - 
can eliminate one mRNA molecule after another, thereby efficiently 
blocking production of the protein that the mRNA encodes.

Two features of miRNAs make them especially useful regulators of gene
expression. First, a single miRNA can regulate a whole set of different
mRNAs so long as the mRNAs carry a common sequence.
Second, a gene that encodes an miRNA occupies relatively little space in
the genome compared with one that encodes a transcription regulator.

\subsection{RNA Interference destroys double-Stranded Foreign RNAs}

targeted RNA degradation mechanism, called RNA interference (RNAi), helps to keep these 
potentially dangerous invaders in check.
The presence of foreign, double-stranded RNA in the cell triggers RNAi by
first attracting a protein complex containing a nuclease called Dicer. Dicer
cleaves the double-stranded RNA into short fragments (approximately 23 nucleotide 
pairs in length) called small interfering RNAs (siRNAs).
These short, double-stranded RNAs are then incorporated into RISCs, the
same complexes that can carry miRNAs. The RISC discards one strand of
the siRNA duplex and uses the remaining single-stranded RNA to locate
a complementary foreign RNA molecule. This target RNA
molecule is then rapidly degraded, leaving the RISC free to search out
more of the same foreign RNA molecules.

\subsection{Scientists Can use RNA Interference to Turn Off Genes}

In a practical sense, RNAi has become a powerful experimental tool that allows scientists to inactivate almost
any gene in cultured cells or, in some cases, a whole plant or animal.

At the same time, RNAi shows real potential as a powerful new approach
for treating human disease. Because many human disorders result from
the inappropriate expression of genes, the ability to turn these genes off
by introducing complementary siRNA molecules holds great medical
promise.

Finally, the discovery that RNAs play such a key role in controlling gene
expression expands our understanding of the types of regulatory networks
that cells have at their command.

\section{Essential concepts}

\begin{itemize}
\item A typical eucaryotic cell expresses only a fraction of its genes, and
the distinct types of cells in multicellular organisms arise because
different sets of genes are expressed as cells differentiate.
\item Although all of the steps involved in expressing a gene can in principle 
be regulated, for most genes the initiation of transcription is the
most important point of control.
\item The transcription of individual genes is switched on and off in cells
by transcription regulators. These act by binding to short stretches of
DNA called regulatory DNA sequences.
\item Although each transcription regulator has unique features, most bind
to DNA using one of a small number of structural motifs. The precise
amino acid sequence that is folded into the DNA-binding motif determines 
the particular DNA sequence that is recognized.
\item In bacteria, transcription regulators usually bind to regulatory DNA
sequences close to where RNA polymerase binds. They can either
activate or repress transcription of the gene. In eucaryotes, regulatory 
DNA sequences are often separated from the promoter by many
thousands of nucleotide pairs.
\item Eucaryotic transcription regulators act in two fundamental ways:
(1) they can directly affect the assembly process of RNA polymerase
and the general transcription factors at the promoter, and (2) they
can locally modify the chromatin structure of promoter regions.
\item In eucaryotes, the expression of a gene is generally controlled by a
combination of transcription regulators.
\item In multicellular plants and animals, the production of different transcription 
regulators in different cell types ensures the expression of
only those genes appropriate to the particular type of cell.
\item Cells in multicellular organisms have mechanisms that enable their
progeny to ‘remember’ what type of cell they should be.
\item A single transcription regulator, if expressed in the appropriate precursor 
cell, can trigger the formation of a specialized cell type or
even an entire organ.
\item Cells can also regulate gene expression by controlling events that
occur after transcription has begun. Many of these mechanisms
rely on RNA molecules that can influence their own transcription or
translation.
\item MicroRNAs (miRNAs) control gene expression by base-pairing with
specific mRNAs and regulating their stability and their translation.
\item Cells have a defense mechanism for destroying ‘foreign’ double-stranded 
RNAs, many of which are produced by viruses. Scientists
can take advantage of this mechanism, called RNA interference, to
inactivate genes of interest by simply injecting cells with double-stranded 
RNAs that target the mRNAs produced by those genes.
\end{itemize}
